# Mental Health

```{r}
#| label: gb
#| include: false
# include does not show output or code #| is a special syntax for Quarto 

#When testing and not rendering the report, the if statement returns true but when rendering as the report, the else executes
if (file.exists("pages/Global/gb.R")) {
  source("pages/Global/gb.R")
  source("pages/Data/data.R")
} else {
  source("./Global/gb.R")
  source("./Data/data.R")
}
```

## Depression

PNM counties and the state of Minnesota show that roughly 1 in every 3 to 5 people is projected to experience depression. This high ratio is also reflected in the Minnesota Student Survey responses from 9th grade students. When asked how often they have felt down, depressed, or hopeless over the past two weeks, about 1 in every 3 to 5 reported feeling this way. Mahnomen County had the highest percentage of 9th graders responding 'yes' to this question.

::: {#fig-conditionDepression}
```{r}
#| label: f_places2022Depression
#| warning: false
#| include: false

# Filter the data by 'measureid'
filtered_data <- f_places2022 |> 
  dplyr::filter(measureid == "DEPRESSION") |> 
  dplyr::mutate(#Passed into ggplot (PDF Plot) but displayed in plotly (HMTL Plot)
    toolTipTextHeader1 = "\nLocation: "
    , toolTipTextValue1 = locationname
    , toolTipTextHeader2 = "\nPrevalence: "
    , toolTipTextValue2 = paste0(formattable::comma(data_value,0), " (", formattable::comma(low_confidence_limit,0), "-", formattable::comma(high_confidence_limit,0), ")")
  )

filtered_dataTbl <- filtered_data |> 
    dplyr::arrange(locationid, year) |> 
    dplyr::mutate(prevalenceType = ifelse(datavaluetypeid == "CrdPrv", "Crude", "Age-Adjusted"),
                  tab = ifelse(datavaluetypeid == "CrdPrv", "Crude Prevalence", "Age-Adjusted Prevalence")) |> 
    dplyr::select(prevalenceType, locationname, low_confidence_limit, data_value, high_confidence_limit, tab)
  
figCap_plotTitle <- paste(filtered_data$measure[1], filtered_data$brfssYr[1])
figSubcap <- "Data Source: @cdc2024Places"
plotXCaption <- paste("Data Source:",cdc2024Places) #Since the x axis is passed into plotly, I am defining the x axis for the plot with the data source so when/if people use the plot in their work the data source is referenced ggplot does a a caption argument but it does not pass into plotly. If the caption should be after the legend, then ggplot caption can be used for the pdf version and annotation can be used in plotly, I'll just have to pass an extra argument into the plotly function.


#Table 
colList <- list("Type", "Location", "Low CI", "Prevalence", "High CI") # Using a list accommodates tables with different column lengths.

#Define the plot here so it is only wrote once
p <- gbFun_plotSingleYrPDF(filtered_data, "locationname", "locationid","data_value", "toolTipTextHeader1", 
    "toolTipTextValue1", "toolTipTextHeader2", "toolTipTextValue2", "data_value_type",
    "low_confidence_limit", "high_confidence_limit", figCap_plotTitle, plotXCaption, "Prevalence",
    gbVal_colorScheme, ifelse(knitr::is_html_output(), 10, 12)
    )
```

<div>

::: {.content-visible when-format="html"}
```{r}
#| fig-width: 8
#| fig-height: 5 

if (knitr::is_html_output()) {
# Create the ggplot plot ggplot and then its wrapped within plotly function
gbFun_plotHTML(p, figCap_plotTitle)
}
```

```{r}
#| tbl-width: 8
#| tbl-height: 4
#| warning: false
#| results: asis
if (knitr::is_html_output()) {
  
  split_data <- split(filtered_dataTbl, filtered_dataTbl$tab)
  
# Generate tabset panels
  cat("::: {.panel-tabset}\n")
     for (tab in names(split_data)) {
       cat(sprintf("## %s\n\n", tab))
       
       # Remove the 'tab' column from the current subset of data
       current_data <- split_data[[tab]][, !names(split_data[[tab]]) %in% "tab"]
       
       print(htmltools::tagList(gbFun_tableHTML(current_data, colList, figCap_plotTitle)))
     }
  cat(":::\n")

}
```
:::

::: {.content-visible when-format="pdf" layout="[[1],[1, 1]]"}
```{r}
#| fig-width: 11
#| fig-height: 3
#| warning: false

if (!knitr::is_html_output()) {
# Plots have to be printed
print(p)
}
```

```{r}
#| tbl-width: 4
#| tbl-height: 3
#| warning: false
if (!knitr::is_html_output()) {
  gbFun_tablePDF(filtered_dataTbl |> dplyr::filter(prevalenceType == "Crude") |> dplyr::select(-tab), colList)
}
```

```{r}
#| tbl-width: 4
#| tbl-height: 3
#| warning: false
if (!knitr::is_html_output()) {
  gbFun_tablePDF(filtered_dataTbl |> dplyr::filter(prevalenceType == "Age-Adjusted") |> dplyr::select(-tab), colList)
}
```
:::

`{r} figSubcap`

</div>

`{r} figCap_plotTitle`
:::

::: {#fig-mssDepression}
```{r}
#| label: f_mss9GradeDepression
#| warning: false
#| include: false

# Data Wrangling so it can just be referenced in the PDF and HTML output instead of writing it twice

filtered_data <- f_mss |> 
  dplyr::filter(mss_topic == "% of 9th Graders reported over the past two weeks: how often have you been bothered: feeling down: depressed or hopeless several days") |> 
  dplyr::mutate(#Passed into ggplot (PDF Plot) but displayed in plotly (HMTL Plot)
    toolTipTextHeader1 = "Year: "
    , toolTipTextValue1 = year
    , toolTipTextHeader2 = "\nLocation: "
    , toolTipTextValue2 = location
    , toolTipTextHeader3 = "\n9th Graders Feeling Depressed: "
    , toolTipTextValue3 = scales::label_percent(accuracy = .01)(pct)
  )

minYr <- unique(min(filtered_data$year))
maxYr <- unique(max(filtered_data$year))

# Filtering data here handles table variations better and is easier to maintain than filtering in the function.

filtered_dataTbl <- filtered_data |> 
  dplyr::filter(year >= maxYr-3)  |>
  dplyr::arrange(fips, year) |>
  dplyr::group_by(fips) |> 
  dplyr::summarize(
    Location = location
    , valueMinYr = pct[year == min(year)]
    , valueMaxYr = pct[year == max(year)]
    , pctDifference = valueMaxYr - valueMinYr
    , .groups = 'drop'
  ) |>
  dplyr::distinct() |> 
  dplyr::select(-fips) |>  
  dplyr::mutate(
    valueMinYr = scales::label_percent(accuracy = .01)(valueMinYr)
    , valueMaxYr = scales::label_percent(accuracy = .01)(valueMaxYr)
    , pctDifference = scales::label_percent(accuracy = .01)(pctDifference)
    ) |>
  dplyr::rename("% Change" = pctDifference) |> 
  dplyr::rename_with(~ paste0("Depressive Feelings ", maxYr-3), valueMinYr)  |> 
  dplyr::rename_with(~ paste0("Depressive Feelings ", maxYr), valueMaxYr)


figCap_plotTitle <- "Percentage of 9th graders reporting over the past two weeks, how often have\nyou been bothered, feeling down, depressed or hopeless several days (MSS)"
figSubcap <- "Data Source: @mdeMinnesotaStudentSurvey"
plotXCaption <- paste("Data Source:", mdeMinnesotaStudentSurvey) #Since the x axis is passed into plotly, I am defining the x axis for the plot with the data source so when/if people use the plot in their work the data source is referenced ggplot does a a caption argument but it does not pass into plotly. If the caption should be after the legend, then ggplot caption can be used for the pdf version and annotation can be used in plotly, I'll just have to pass an extra argument into the plotly function.

# Plot
scaleYContinuous = scales::label_percent()  # Format y-axis values with geography
#Table
colList <- colnames(filtered_dataTbl) # Using a list accommodates tables with different column lengths.

#Define the plot here so it is only wrote once
p <- gbFun_plotTrendPDF(
    filtered_data, "year", "pct", "location", "fips", "toolTipTextHeader1", "toolTipTextValue1", 
    "toolTipTextHeader2", "toolTipTextValue2", "toolTipTextHeader3", "toolTipTextValue3", figCap_plotTitle,
    plotXCaption, "", minYr, maxYr, scaleYContinuous,
    gbVal_colorScheme, gbVal_shapeScheme, ifelse(knitr::is_html_output(), 10, 14)
    )

```

<div>

::: {.content-visible when-format="html"}
```{r}
#| fig-width: 8
#| fig-height: 5
#| warning: false
if (knitr::is_html_output()) {
  # Create the ggplot plot ggplot and then its wrapped within plotly function
  gbFun_plotHTML(p, figCap_plotTitle)
}
```

```{r}
#| tbl-width: 8
#| tbl-height: 4
#| warning: false
if (knitr::is_html_output()) {
  gbFun_tableHTML(filtered_dataTbl, colList, figCap_plotTitle)
}
```
:::

::: {.content-visible when-format="pdf" layout="[[1, 1], [1]]"}
```{r}
#| warning: false
#| fig-width: 9
#| fig-height: 5
if (!knitr::is_html_output()) {
# Plots have to be printed
print(p)
}
```

```{r}
#| warning: false
#| tbl-width: 2
#| tbl-height: 5
if (!knitr::is_html_output()) {
#Tables shouldn't be printed because the HTML table will display
gbFun_tablePDF(filtered_dataTbl, colList)
}
```
:::

`{r} figSubcap`

</div>

`{r} figCap_plotTitle`
:::

## Optimal Care for Depression

A very encouraging sign is Polk, Norman, and Mahnomen counties appear to be doing well in screening for mental health and depression for patients 12-17 years of age. PNM Counties are slightly lower than Minnesota but all three counties are over 80%. Early screening plays a crucial role in identifying and addressing mental health issues promptly, leading to better outcomes for our youth.

::: {#fig-pufAmh}
```{r}
#| label: f_pufAmh
#| warning: false
#| include: false

# Data Wrangling so it can just be referenced in the PDF and HTML output instead of writing it twice

filtered_data <- f_puf |> 
  dplyr::filter(measure == "amh") |> 
  dplyr::mutate(#Passed into ggplot (PDF Plot) but displayed in plotly (HMTL Plot)
    toolTipTextHeader1 = "Year: "
    , toolTipTextValue1 = year
    , toolTipTextHeader2 = "\nLocation: "
    , toolTipTextValue2 = location
    , toolTipTextHeader3 = "\nOptimal Care Rate: "
    , toolTipTextValue3 =  scales::label_number(accuracy = 0.01)(optimal_care_rate)
  )

minYr <- unique(min(filtered_data$year))
maxYr <- unique(max(filtered_data$year))

# Filtering data here handles table variations better and is easier to maintain than filtering in the function.
filtered_dataTbl <- filtered_data |> 
  dplyr::filter(year == maxYr)  |>
  dplyr::arrange(fips, year) |>
  dplyr::mutate(placeholder = scales::label_number(accuracy = 0.01)(optimal_care_rate)) |>   
  dplyr::select(year, location, placeholder) # Ensure fields are ordered as desired.  


figCap_plotTitle <- "Adolescent Mental Health and/or Depression Screening" 
figSubcap <- "Data Source: @mdhPublicUseFiles2022"
plotXCaption <- paste("Data Source:", mdhPublicUseFiles2022) #Since the x axis is passed into plotly, I am defining the x axis for the plot with the data source so when/if people use the plot in their work the data source is referenced ggplot does a a caption argument but it does not pass into plotly. If the caption should be after the legend, then ggplot caption can be used for the pdf version and annotation can be used in plotly, I'll just have to pass an extra argument into the plotly function.

# Plot
scaleYContinuous =  scales::label_number(accuracy = 0.01)  # Format y-axis values with geography
#Table
colList <- list("Year", "Location", "Optimal Care Rate") # Using a list accommodates tables with different column lengths.


#Define the plot here so it is only wrote once
p <- gbFun_plotTrendPDF(
    filtered_data, "year", "optimal_care_rate", "location", "fips", "toolTipTextHeader1", 
    "toolTipTextValue1", "toolTipTextHeader2", "toolTipTextValue2", "toolTipTextHeader3",
    "toolTipTextValue3", figCap_plotTitle, plotXCaption, "", minYr, maxYr, scaleYContinuous,
    gbVal_colorScheme, gbVal_shapeScheme, ifelse(knitr::is_html_output(), 10, 14)
    )
```

<div>

::: {.content-visible when-format="html"}
```{r}
#| fig-width: 8
#| fig-height: 5
#| warning: false
if (knitr::is_html_output()) {
  # Create the ggplot plot ggplot and then its wrapped within plotly function
  gbFun_plotHTML(p, figCap_plotTitle)
}
```

```{r}
#| tbl-width: 8
#| tbl-height: 4
#| warning: false
if (knitr::is_html_output()) {
  gbFun_tableHTML(filtered_dataTbl, colList, figCap_plotTitle)
}
```
:::

::: {.content-visible when-format="pdf" layout="[[1, 1], [1]]"}
```{r}
#| warning: false
#| fig-width: 9
#| fig-height: 5
if (!knitr::is_html_output()) {
# Plots have to be printed
print(p)
}
```

```{r}
#| warning: false
#| tbl-width: 2
#| tbl-height: 5
if (!knitr::is_html_output()) {
#Tables shouldn't be printed because the HTML table will display
gbFun_tablePDF(filtered_dataTbl, colList)
}
```
:::

`{r} figSubcap`

</div>

`{r} figCap_plotTitle`
:::

## Community Support/ Risky Behavior

In terms of community support, Mahnomen County saw a positive change of 12.3% in students feeling that the community cared for them "quite a bit" or "very much." Conversely, Polk County experienced a 13.5% decrease in this sentiment.

The Minnesota Student Survey helps identify early signs of possible problematic behavior in our youth. For instance, we can see that 60.8% of 11th graders in Polk County reported not using alcohol, marijuana, or drugs within the last year. Unfortunately, we currently lack data on how well Mahnomen and Norman counties are doing in this area.

::: {#fig-mssCommunityCare}
```{r}
#| label: f_mss9GradeComunityCare
#| warning: false
#| include: false

# Data Wrangling so it can just be referenced in the PDF and HTML output instead of writing it twice

filtered_data <- f_mss |> 
  dplyr::filter(mss_topic == "% of 9th Graders community cared about them \"quite a bit\" or \"very much\"") |> 
  dplyr::mutate(#Passed into ggplot (PDF Plot) but displayed in plotly (HMTL Plot)
    toolTipTextHeader1 = "Year: "
    , toolTipTextValue1 = year
    , toolTipTextHeader2 = "\nLocation: "
    , toolTipTextValue2 = location
    , toolTipTextHeader3 = "\n9th Graders Community Cared: "
    , toolTipTextValue3 = scales::label_percent(accuracy = .01)(pct)
  )

minYr <- unique(min(filtered_data$year))
maxYr <- unique(max(filtered_data$year))

# Filtering data here handles table variations better and is easier to maintain than filtering in the function.
filtered_dataTbl <- filtered_data |> 
  dplyr::filter(year >= maxYr-3)  |>
  dplyr::arrange(fips, year) |>
  dplyr::group_by(fips) |> 
  dplyr::summarize(
    Location = location
    , valueMinYr = pct[year == min(year)]
    , valueMaxYr = pct[year == max(year)]
    , pctDifference = valueMaxYr - valueMinYr
    , .groups = 'drop'
  ) |>
  dplyr::distinct() |> 
  dplyr::select(-fips) |>  
  dplyr::mutate(
    valueMinYr = scales::label_percent(accuracy = .01)(valueMinYr)
    , valueMaxYr = scales::label_percent(accuracy = .01)(valueMaxYr)
    , pctDifference = scales::label_percent(accuracy = .01)(pctDifference)
    ) |>
  dplyr::rename("% Change" = pctDifference) |> 
  dplyr::rename_with(~ paste0("Community Cared ", maxYr-3), valueMinYr)  |> 
  dplyr::rename_with(~ paste0("Community Cared ", maxYr), valueMaxYr) 


figCap_plotTitle <- "Percentage of 9th graders who reported that\nthe community cared about them \"quite a bit\" or \"very much\" (MSS)"
figSubcap <- "Data Source: @mdeMinnesotaStudentSurvey"
plotXCaption <- paste("Data Source:", mdeMinnesotaStudentSurvey) #Since the x axis is passed into plotly, I am defining the x axis for the plot with the data source so when/if people use the plot in their work the data source is referenced ggplot does a a caption argument but it does not pass into plotly. If the caption should be after the legend, then ggplot caption can be used for the pdf version and annotation can be used in plotly, I'll just have to pass an extra argument into the plotly function.

# Plot
scaleYContinuous = scales::label_percent()  # Format y-axis values with geography
#Table
colList <- colnames(filtered_dataTbl) # Using a list accommodates tables with different column lengths.

#Define the plot here so it is only wrote once
p <- gbFun_plotTrendPDF(
    filtered_data, "year", "pct", "location", "fips", "toolTipTextHeader1", "toolTipTextValue1", 
    "toolTipTextHeader2", "toolTipTextValue2", "toolTipTextHeader3", "toolTipTextValue3", figCap_plotTitle,
    plotXCaption, "", minYr, maxYr, scaleYContinuous,
    gbVal_colorScheme, gbVal_shapeScheme, ifelse(knitr::is_html_output(), 10, 14)
    )
```

<div>

::: {.content-visible when-format="html"}
```{r}
#| fig-width: 8
#| fig-height: 5
#| warning: false
if (knitr::is_html_output()) {
  # Create the ggplot plot ggplot and then its wrapped within plotly function
  gbFun_plotHTML(p, figCap_plotTitle)
}
```

```{r}
#| tbl-width: 8
#| tbl-height: 4
#| warning: false
if (knitr::is_html_output()) {
  gbFun_tableHTML(filtered_dataTbl, colList, figCap_plotTitle)
}
```
:::

::: {.content-visible when-format="pdf" layout="[[1, 1], [1]]"}
```{r}
#| warning: false
#| fig-width: 9
#| fig-height: 5
if (!knitr::is_html_output()) {
# Plots have to be printed
print(p)
}
```

```{r}
#| warning: false
#| tbl-width: 2
#| tbl-height: 5
if (!knitr::is_html_output()) {
#Tables shouldn't be printed because the HTML table will display
gbFun_tablePDF(filtered_dataTbl, colList)
}
```
:::

`{r} figSubcap`

</div>

`{r} figCap_plotTitle`
:::

::: {#fig-mssNoDrugUse}
```{r}
#| label: f_mss11GradeNoDrugUse
#| warning: false
#| include: false

# Data Wrangling so it can just be referenced in the PDF and HTML output instead of writing it twice

filtered_data <- f_mss |> 
  dplyr::filter(mss_topic == "% of 11th graders in the PNM service area who have not used alcohol: marijuana: and/or drugs in the past year") |> 
  dplyr::mutate(#Passed into ggplot (PDF Plot) but displayed in plotly (HMTL Plot)
    toolTipTextHeader1 = "Year: "
    , toolTipTextValue1 = year
    , toolTipTextHeader2 = "\nLocation: "
    , toolTipTextValue2 = location
    , toolTipTextHeader3 = "\n11th Graders No Substance Use: "
    , toolTipTextValue3 = scales::label_percent(accuracy = .01)(pct)
  )

minYr <- unique(min(filtered_data$year))
maxYr <- unique(max(filtered_data$year))

# Filtering data here handles table variations better and is easier to maintain than filtering in the function.
filtered_dataTbl <- filtered_data |> 
  dplyr::filter(year >= maxYr-3)  |>
  dplyr::arrange(fips, year) |>
  dplyr::group_by(fips) |> 
  dplyr::summarize(
    Location = location
    , valueMinYr = pct[year == min(year)]
    , valueMaxYr = pct[year == max(year)]
    , pctDifference = valueMaxYr - valueMinYr
    , .groups = 'drop'
  ) |>
  dplyr::distinct() |> 
  dplyr::select(-fips) |>  
  dplyr::mutate(
    valueMinYr = scales::label_percent(accuracy = .01)(valueMinYr)
    , valueMaxYr = scales::label_percent(accuracy = .01)(valueMaxYr)
    , pctDifference = scales::label_percent(accuracy = .01)(pctDifference)
    ) |>
  dplyr::rename("% Change" = pctDifference) |> 
  dplyr::rename_with(~ paste0("No Substance Use ", maxYr-3), valueMinYr)  |> 
  dplyr::rename_with(~ paste0("No Substance Use ", maxYr), valueMaxYr)   


figCap_plotTitle <- "Percentage of 11th graders who have not used\nalcohol, marijuana, and/or drugs in the past year (MSS)"
figSubcap <- "Data Source: @mdeMinnesotaStudentSurvey"
plotXCaption <- paste("Data Source:", mdeMinnesotaStudentSurvey) #Since the x axis is passed into plotly, I am defining the x axis for the plot with the data source so when/if people use the plot in their work the data source is referenced ggplot does a a caption argument but it does not pass into plotly. If the caption should be after the legend, then ggplot caption can be used for the pdf version and annotation can be used in plotly, I'll just have to pass an extra argument into the plotly function.

# Plot
scaleYContinuous = scales::label_percent()  # Format y-axis values with geography
#Table
colList <- colnames(filtered_dataTbl) # Using a list accommodates tables with different column lengths.

#Define the plot here so it is only wrote once
p <- gbFun_plotTrendPDF(
    filtered_data, "year", "pct", "location", "fips", "toolTipTextHeader1", "toolTipTextValue1", 
    "toolTipTextHeader2", "toolTipTextValue2", "toolTipTextHeader3", "toolTipTextValue3", figCap_plotTitle,
    plotXCaption, "", minYr, maxYr, scaleYContinuous,
    gbVal_colorScheme, gbVal_shapeScheme, ifelse(knitr::is_html_output(), 10, 14)
    )
```

<div>

::: {.content-visible when-format="html"}
```{r}
#| fig-width: 8
#| fig-height: 5
#| warning: false
if (knitr::is_html_output()) {
  # Create the ggplot plot ggplot and then its wrapped within plotly function
  gbFun_plotHTML(p, figCap_plotTitle)
}
```

```{r}
#| tbl-width: 8
#| tbl-height: 4
#| warning: false
if (knitr::is_html_output()) {
  gbFun_tableHTML(filtered_dataTbl, colList, figCap_plotTitle)
}
```
:::

::: {.content-visible when-format="pdf" layout="[[1, 1], [1]]"}
```{r}
#| warning: false
#| fig-width: 9
#| fig-height: 5
if (!knitr::is_html_output()) {
# Plots have to be printed
print(p)
}
```

```{r}
#| warning: false
#| tbl-width: 2
#| tbl-height: 5
if (!knitr::is_html_output()) {
#Tables shouldn't be printed because the HTML table will display
gbFun_tablePDF(filtered_dataTbl, colList)
}
```
:::

`{r} figSubcap`

</div>

`{r} figCap_plotTitle`
:::

## Suicide

Suicide is complex; there is no single cause of death by suicide. Suicide prevention efforts are based on evidence that most suicides are preventable, mental illness is treatable, and recovery is possible.

In 2022, 860 Minnesotans died from suicide, the highest total ever, and preliminary data indicates 815 died in 2023. This translates to an age-adjusted rate of 14.8 and 14.1 per 100,000, respectively. Preliminary figures for 2023 are based on projections made using Minnesota death certificates finalized as of May 1, 2024. Continuing a persistent trend, males had a higher suicide rate than females in 2022 and 2023. In 2022 and 2023, American Indian or Alaska Natives had a higher suicide rate than the most populous racial or ethnic groups in Minnesota. The largest percentage of suicide deaths in 2023, 47%, involved a firearm. Like previous years, suicide rates in rural areas in 2023 tended to be higher than urban areas. (Data Brief: Suicide Up in 2022, Down in 2023 In 2022 (2024), Minnesota Department of Health) [Suicide 2021 data brief](https://www.health.state.mn.us/communities/suicide/documents/2023suicidedatabrief.pdf)

From 2016-2021, there were fifty reported deaths by suicide in Polk, Norman and Mahnomen counties -- making up 48% of the total deaths by suicide for the eight Northwest Minnesota county region. Firearms were the most common mechanism of injury accounting for suicide deaths. (Minnesota Department of Health)

In addition to suicide deaths, data on suicide ideation and self-harm provides important insight into the broader mental health challenges facing communities. Hospital and emergency room visits for self-harm are a key indicator of emotional distress and potential suicide risk. In 2023, Mahnomen County had the highest age-adjusted rate of self-harm hospital and ER visits at 537.3 per 100,000, an increase from 519.8 in 2022. In contrast, Minnesota as a whole saw a decrease from 208.5 to 193.1, and both Norman and Polk counties also experienced declines. These patterns suggest that while some areas are seeing improvement, others continue to face rising mental health concerns. Monitoring self-harm trends helps identify where prevention and support services are most needed, especially for youth and vulnerable populations.

::: {#fig-suicideIdeaitionSelfHarm}
```{r}
#| label: f_suicideIdeaitionSelfHarm
#| warning: false
#| include: false

# Data Wrangling so it can just be referenced in the PDF and HTML output instead of writing it twice

filtered_data <- f_suicideIdeationSelfHarm |> 
  dplyr::mutate(#Passed into ggplot (PDF Plot) but displayed in plotly (HMTL Plot)
    toolTipTextHeader1 = "Year: "
    , toolTipTextValue1 = year
    , toolTipTextHeader2 = "\nLocation: "
    , toolTipTextValue2 = location
    , toolTipTextHeader3 = "\nAge Adjusted Suicide Ideation/Self Harm: "
    , toolTipTextValue3 = scales::label_comma(accuracy = .01)(ageAdjustedRate)
  )

minYr <- unique(min(filtered_data$year))
maxYr <- unique(max(filtered_data$year))
# Filtering data here handles table variations better and is easier to maintain than filtering in the function.
filtered_dataTbl <- filtered_data |> 
  dplyr::filter(year >= maxYr-1)  |>
  dplyr::arrange(fips, year) |>
  dplyr::group_by(fips) |> 
  dplyr::summarize(
    Location = location
    , valueMinYr = ageAdjustedRate[year == min(year)]
    , valueMaxYr = ageAdjustedRate[year == max(year)]
    , pctDifference = valueMaxYr - valueMinYr
    , .groups = 'drop'
  ) |>
  dplyr::distinct() |> 
  dplyr::select(-fips) |>  
  dplyr::mutate(
    valueMinYr = scales::label_comma(accuracy = .01)(valueMinYr)
    , valueMaxYr = scales::label_comma(accuracy = .01)(valueMaxYr)
    , pctDifference = scales::label_comma(accuracy = .01)(pctDifference)
    ) |>
  dplyr::rename("Rate Change" = pctDifference) |> 
  dplyr::rename_with(~ paste0("Age-Adjusted Rate ", maxYr-1), valueMinYr)  |> 
  dplyr::rename_with(~ paste0("Age-Adjusted Rate ", maxYr), valueMaxYr)


figCap_plotTitle <- "Age-Adjusted Rate of Self-Harm Hospital/ER Visits (per 100,000)"
figSubcap <- "Data Source: @mdhSuicideIdeationSelfHarm"
plotXCaption <- paste("Data Source:", mdhSuicideIdeationSelfHarm) #Since the x axis is passed into plotly, I am defining the x axis for the plot with the data source so when/if people use the plot in their work the data source is referenced ggplot does a a caption argument but it does not pass into plotly. If the caption should be after the legend, then ggplot caption can be used for the pdf version and annotation can be used in plotly, I'll just have to pass an extra argument into the plotly function.

# Plot
scaleYContinuous = scales::label_comma()  # Format y-axis values with geography
#Table
colList <- colnames(filtered_dataTbl) # Using a list accommodates tables with different column lengths.

#Define the plot here so it is only wrote once
p <- gbFun_plotTrendPDF(
    filtered_data, "year", "ageAdjustedRate", "location", "fips", "toolTipTextHeader1", "toolTipTextValue1", 
    "toolTipTextHeader2", "toolTipTextValue2", "toolTipTextHeader3", "toolTipTextValue3", figCap_plotTitle,
    plotXCaption, "", minYr, maxYr, scaleYContinuous,
    gbVal_colorScheme, gbVal_shapeScheme, ifelse(knitr::is_html_output(), 10, 14)
    )

```

<div>

::: {.content-visible when-format="html"}
```{r}
#| fig-width: 8
#| fig-height: 5
#| warning: false
if (knitr::is_html_output()) {
  # Create the ggplot plot ggplot and then its wrapped within plotly function
  gbFun_plotHTML(p, figCap_plotTitle)
}
```

```{r}
#| tbl-width: 8
#| tbl-height: 4
#| warning: false
if (knitr::is_html_output()) {
  gbFun_tableHTML(filtered_dataTbl, colList, figCap_plotTitle)
}
```
:::

::: {.content-visible when-format="pdf" layout="[[1, 1], [1]]"}
```{r}
#| warning: false
#| fig-width: 9
#| fig-height: 5
if (!knitr::is_html_output()) {
# Plots have to be printed
print(p)
}
```

```{r}
#| warning: false
#| tbl-width: 2
#| tbl-height: 5
if (!knitr::is_html_output()) {
#Tables shouldn't be printed because the HTML table will display
gbFun_tablePDF(filtered_dataTbl, colList)
}
```
:::

`{r} figSubcap`

</div>

`{r} figCap_plotTitle`
:::
